Insert the article title here

### 2021-07-01 fstats_pca_lewontin.tex
discussion

\section{$F$-statistics in PCA-space}
As shown by e.g. \cite{oteo-garcia2021}, $F$-statistics can be thought of as inner products in Euclidean space, and $F_2$ is an (estimated) squared Euclidean distance between two populations in allele frequency space. By performing a PCA, we just translate and rotate our data, but Euclidean distances and dot products are both invariant under both these operations. Hence, neither mean-centering (a translation) nor PCA (a rotation) will change $F_2$. What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or the principal components $\MP$. Formally,

@@

\section{Results}
\subsection{$F_4$}
Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample. In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.

@@

@@

\section{Discussion}
\begin{itemize}
    \item weighting of SNPs
    \item estimation error
    \item propagating erros
    \item exploratory data analysis
    \item missing data
    \item population vs. sample allele frequencies
\end{itemize}

>>>>

\section{$F$-statistics in PCA-space}
As shown by e.g. \cite{oteo-garcia2021}, $F$-statistics can be thought of as inner products in Euclidean space, and $F_2$ is an (estimated) squared Euclidean distance between two populations in allele frequency space. By performing a PCA, we just translate and rotate our data, but Euclidean distances and dot products are both invariant under both these operations. Hence, neither mean-centering (a translation) nor PCA (a rotation) will change $F_2$. What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or any other orthogonal basis such as the principal components $\MP$. Formally,

@@

@@

\subsection{$F_4$}
Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample. In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.

@@

\section{Discussion}
Particularly for the analysis of ancient DNA, $F$-statistics have been established as a powerful tool to describe population genetic diversity, but they have a number of limitations. In particular, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \cite{patterson2012,harney2020}. As a consequence, researchers concerned about model fits may ascertain reference populations in a way that satisfies these assumptions, thus inadvertently making population structure appear sparser than it truly is. This is perhaps most obvious from Figure \ref{fig:f3}B, where large gaps are present. However, these gaps are due to sparse sampling that disappear if more populations were sampled \cite[e.g.][]{peter2020a}, not due to gaps in genetic diversity. If population ascertainment is not done very carefully, tools built on top of $F$-statistics, such as \texttt{qpGraph} and \texttt{qpAdm}, may thus only provide a very loose lower bound for the number of gene flow events.

In contrast, the perspective on $F$-statistics in data space \citep{oteo-garcia2021} and on PCA does not require assumptions on number of admixture or gene flow events. Independent of any model, A population $X_x$ can be thought of as admixed between $ X_1$ and $X_2$ if it lies in the ball with diameter $\overline{X_1X_2}$. And independent of any models, two axes of variation that are not perpendicular to each other shows that there is some degree of shared history.

The connection between the graph-based and PCA-based interpretations of $F_3$ and $F_4$ is due to concept of projection. As illustrated above, both $F$-statistics can be thought of as projections on any particular axis of variation. If that variation corresponds to a tree-branch, the interpretations align. A collorary to this interpretation is the importance of orthogonality, or independence for describing gene flow. Populations without gene-flow will evolve independently, and their changes in allele frequency will therefore be orthogonal, resulting in $F_4$-statistics of zero, and right angles on a PCA plot. Even though this is only true when considering all PCs, the result holds reasonably well when just looking at the first two PCs -- in our example of the world-overview data, I found that the variation in Sub-Saharan Africa is mostly -- but not completely -- uncorrelated with the variation between European and Asian populations (Figure \ref{fig:f4}A).

While the data space produces a much larger model space than trees, the main drawback of the PCA-based interpretation is a lack of interpretability; it is not easy to define a generative model that could generate a complex PCA-plot.

 However, despite the apparent complexity, the PC-spaces \emph{are} sparse, and using just ten or even two PCs often gives very good approximations.

To make PCA and $F$-statistics more comparable in practical settings, there are a number of -- mainly statistical -- concerns that still need to be addressed in future work. The perhaps most obvious one is that PCA is most frequently run on individuals, whereas $F$-statistics are often calculated on populations. This is not a conceptual issue, as both PCA and $F$-statistics can be run on either \cite{cavalli-sforza1994}. Population based analyses have the advantage that they are easier to interpret and compute (current packages are ill-equipped to calculate all pairwise $F$-statistics between data sets with thousands of individuals \cite{patterson2012}). However, this requires the assumption that the within-population variation is independent from the between-population variation; something that is analogous to the variance partitioning based on PCs here.

A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \cite{patterson2006}. This is only a difference of convention; $F$-statistics could also be calculated using the same weighting. The close connection between the two approaches developed here suggest that for most analyses, users might want to be consistent and use the same weighting for both types of analyses.

The third and perhaps biggest gap are statistical issues. The treatment here focusses on the mean estimated $F$-statistic, but many  applications of $F$-statistics are based on hypothesis tests \cite{patterson2012}. This requires estimating accurate standard errors for these statistics, which is difficult since nearby SNPs will be correlated \cite{hahn2019}. In contrast, standard PCA does not modeljointly models the covariance matrix due to population structure and sampling. On the other hand, for both data sets I investigated here, the matrix $\MF$ of $F$-statistics estimated using admixtools2 is not a proper squared Euclidean distance matrix, i.e. it is not negative semidefinite and has imaginary PCs. This is not a practical when considering single $F$-statistics or PCA (for analyses here, I used a nearby matrix  \citep{higham2002} with no apparent loss of precision). It does however mean that tools that use matrices of $F$-statistics, such as \texttt{qpadm} or \texttt{qpgraph} may be ill-calibrated, which may partly explain why they generally have poor out-of-sample predictive power and are restricted to a few dozen samples at a time. A model-based framework based on probabilistic PCA \citep{meisner2021, agrawal2020, softimpute} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.

\begin{itemize}
    \item weighting of SNPs
    \item estimation error
    \item propagating erros
    \item exploratory data analysis
    \item missing data
    \item population vs. sample allele frequencies
\end{itemize}
